---
title: "Forest Fire Simulation"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview

This is a demo of the **Drossel–Schwabl forest-fire model**, a simple
cellular automaton that produces self-organized criticality. Each cell in a
grid is in one of three states:

- **empty** (0)
- **tree** (1)
- **burning** (2)

At every time step, the following rules are applied simultaneously to every
cell:

1. A burning cell becomes empty.
2. A tree with at least one burning neighbor catches fire.
3. A tree with no burning neighbor catches fire anyway with probability `f`
   (lightning strike).
4. An empty cell grows a tree with probability `p`.

Only base R is used, so this runs with no extra packages beyond what's in
`renv.lock`.

## Simulation function

```{r simulate-function}
simulate_forest <- function(nrow, ncol, steps, p, f, seed = 1) {
  set.seed(seed)
  grid <- matrix(0L, nrow, ncol)
  history <- vector("list", steps)
  tree_frac <- numeric(steps)
  burning_frac <- numeric(steps)

  for (t in seq_len(steps)) {
    burning <- grid == 2L

    # 4-neighbor burning check via a zero-padded border (fixed boundary)
    pad <- matrix(FALSE, nrow + 2, ncol + 2)
    pad[2:(nrow + 1), 2:(ncol + 1)] <- burning
    neighbor_burning <-
      pad[1:nrow,       2:(ncol + 1)] |
      pad[3:(nrow + 2),  2:(ncol + 1)] |
      pad[2:(nrow + 1), 1:ncol] |
      pad[2:(nrow + 1), 3:(ncol + 2)]

    trees <- grid == 1L
    ignite_neighbor <- trees & neighbor_burning
    ignite_lightning <- trees & !ignite_neighbor &
      (matrix(runif(nrow * ncol), nrow, ncol) < f)

    new_grid <- grid
    new_grid[burning] <- 0L
    new_grid[ignite_neighbor | ignite_lightning] <- 2L

    empty_now <- grid == 0L
    grow <- empty_now & (matrix(runif(nrow * ncol), nrow, ncol) < p)
    new_grid[grow] <- 1L

    grid <- new_grid
    history[[t]] <- grid
    tree_frac[t] <- mean(grid == 1L)
    burning_frac[t] <- mean(grid == 2L)
  }

  list(history = history, tree_frac = tree_frac,
       burning_frac = burning_frac, final = grid)
}
```

## Run the simulation

```{r run-simulation}
nrow_grid <- 60
ncol_grid <- 60
steps <- 200
p_growth <- 0.06     # probability an empty cell grows a tree
f_lightning <- 0.005 # probability a tree ignites spontaneously

result <- simulate_forest(nrow_grid, ncol_grid, steps, p_growth, f_lightning)
```

## Snapshots over time

```{r snapshots, fig.width=9, fig.height=6}
plot_grid <- function(g, main) {
  cols <- c("white", "forestgreen", "orangered")
  image(t(apply(g, 2, rev)), col = cols, breaks = c(-0.5, 0.5, 1.5, 2.5),
        axes = FALSE, main = main, asp = 1)
}

snap_steps <- round(seq(1, steps, length.out = 6))
par(mfrow = c(2, 3), mar = c(1, 1, 2, 1))
for (s in snap_steps) {
  plot_grid(result$history[[s]], main = paste("step", s))
}
```

White = empty, green = tree, orange = burning.

## Forest cover and fires over time

```{r timeseries, fig.width=8, fig.height=4}
par(mfrow = c(1, 1), mar = c(4, 4, 2, 1))
plot(result$tree_frac, type = "l", col = "forestgreen", lwd = 2,
     xlab = "time step", ylab = "fraction of grid",
     main = "Tree cover vs. fire activity", ylim = c(0, 1))
lines(result$burning_frac * 10, col = "orangered", lwd = 1)
legend("topright", legend = c("tree cover", "burning (x10, for visibility)"),
       col = c("forestgreen", "orangered"), lwd = 2, bty = "n")
```

After an initial transient, tree cover fluctuates around a dynamic
equilibrium: growth (`p`) slowly rebuilds the forest until a fire — sparked
either by lightning or by spreading through a connected patch of trees —
burns part of it back down. This growth/burn balance, and the wide range of
fire sizes it produces, is the classic signature of self-organized
criticality in this model.

## Try it yourself

Adjust `p_growth` and `f_lightning` in the **Run the simulation** chunk and
re-knit:

- Increasing `p_growth` relative to `f_lightning` lets denser forests build
  up before burning, producing larger, rarer fires.
- Increasing `f_lightning` produces smaller, more frequent fires and lower
  average tree cover.
